\chapter{Conclusions} \label{chap:concl}

    Collaboration is an integral part of human life, and it is essential to understand how to effectively enable it in digital tools to ensure social information is preserved and users remain aware of it. Virtual Reality (VR) holds great promise in this regard, as it facilitates more natural interactions with the digital world and allows users to share the same space. However, prolonged VR use can be physically tiring, limiting user engagement duration. DeskVR addresses this issue by allowing users to remain seated at their desks while fully immersed in a virtual environment, thus reducing the physical strain of VR use. This dissertation aimed to explore and design a collaborative approach that enables seated users to interact with a virtual environment, considering the constraints of a seated position, and to evaluate how this approach impacts user experience and collaboration.

    The literature review explored concepts of social and workspace awareness, concurrency control, and existing work in the field of DeskVR. Although the study on concurrency control did not suggest specific avenues for exploration, it introduced the idea of personal workspaces, leading to an investigation of the world-in-miniature concept. Research in DeskVR highlighted the value of touch-based interactions in enhancing user comfort and reducing the physical strain of VR. The study on social and workspace awareness showed the importance of these concepts in collaboration and identified their key components. With this understanding, the requirements for the proposed approach were defined: to enable users to communicate collaboratively and effectively about objects or areas of interest in 3D models while minimizing physical effort and providing workspace awareness, all while remaining seated.

    This dissertation proposes Replico, a collaborative approach for DeskVR that allows users to communicate about 3D models in a virtual environment using the world-in-miniature metaphor. Replico enables users to explore the 3D model using touch controls to manipulate a personalized miniature replica and to create points of interest using the Balloon Selection metaphor. These points of interest are replicated in both the miniature and the 3D model. Users are anchored to virtual tables representing their real-life counterparts and can join other users' tables to share their perspectives. Additionally, users can teleport around the 3D model to explore it from different angles in true scale. The miniature displays social information about users, such as their positions, and identifies them by appearance. Points of interest are appearance-coded to correspond with the user who created them.

    A prototype was developed using Unity to evaluate the proposed approach. Gesture detection was implemented using a state machine and uses the K-means clustering algorithm to track the user's hands. The system tracked the user's table with a VR controller to align with the real-life table. Various forms of visual feedback, such as finger trails and touch frame limits, were provided to enhance user interaction. Networking functionality was implemented using Unity's Netcode for GameObjects.

    The approach was evaluated through a user study with 20 participants, forming ten pairs. Participants were required to perform five tasks to assess the efficacy and efficiency of several Replico techniques. These tasks were conducted in two scenarios to test the approach's applicability to various 3D models: a large city and a small but detailed Mars Perseverance rover. After each scenario, participants filled out forms based on the NASA-TLX to evaluate the usability of the approach and their experience with it.

    The results showed that participants could efficiently create points of interest in both scenarios, with no significant differences in task completion and active time. Replico effectively notified users of new points of interest, though larger models required more effort from users. The world-in-miniature metaphor proved useful for communicating points of interest, especially in the smaller rover scenario. While users generally found the approach user-friendly and not physically demanding, they struggled with the teleportation gesture, indicating a need for a more intuitive solution.

    These findings suggest that the approach is practical for collaborative interaction in VR, enabling efficient and user-friendly creation and acknowledgment of points of interest. The study highlights the potential and versatility of the world-in-miniature metaphor for enhancing workspace awareness and social information sharing in large, complex environments and smaller, detailed models. Additionally, it demonstrates the value of touch-based interactions and DeskVR in reducing physical strain and improving user comfort. However, the complexity of the environment increases the effort required from users, and the teleportation gesture needs further refinement. These insights point to areas for future improvement and further development.

\section{Future Work}

    The evaluation of Replico revealed several areas for improvement and avenues for future work. One key area for improvement is the teleportation gesture, which was found to be confusing and difficult to execute. Future work could explore alternative gestures or methods for teleportation that are more intuitive and user-friendly. The teleportation orientation gesture could also be refined. A research article on desigining tactile interfaces for older users \cite{elderly} suggests that touches should not have multiple functions based on how long they are held or how quickly they are pressed, or even double taps. This could be applied to the teleportation gesture, which currently uses a long press to initiate the teleportation. Another improvement that can help with the teleportation gesture is to make the finger tracking more robust to hardware limitations, or to make the gestures more resilient to faulty tracking and accidental touches. A step-by-step interactive guide could assist users in learning the gestures, and a help menu could be added to the system to provide users with information on how to use the system.
    
    The feedback on gestures can be improved. For example color-coding the finger trails to provide feedback on the user's actions, like the visual feedback provided in the implementation of SIT6 \cite{almeidaSIT6IndirectTouchbased2023}. For example, fingers from one hand could be red, while fingers from the other hand could be blue. Another example is to provide audio feedback when a gesture is recognized, a finger is detected, or a point of interest is created, like suggested in a study about how auditory feedback can improve awareness in tabletop interfaces \cite{auditoryAwareness}. Yet another form of feedback could be haptic feedback, such as the use of vibrotactile feedback for when a point of interest is created.

    Additionally, the teleportation behavior itself could be improved by matching the user's head to the balloon's position for a better frame or reference to where the user will be teleported. As it stands, its ambiguous where the vertical position of the user will be after teleportation.

    Another point of improvement is to address the issue of points of interest obscuring the objects they were meant to call attention to. Future work could explore ways to make points of interest more transparent or to provide users with the ability to adjust their transparency.

    Users also reported difficulty in determining when they were within the touch frame limits. Future work could explore ways to provide clearer feedback on the touch frame limits, such as fading in from either inside or outside the frame limits. This requires a more complex shader implementation, but it could be a valuable addition to the system.

    New interactions could be added to the approach to enhance collaboration and communication. For example, users could be able to create zones of interest, encompassing an area instead of a single point. Users could also be able to delete and acknowledge multiple points at once, by increasing the size of the balloon. This could also be used to create larger or smaller points of interest, to give priority to some points. More transformations could be added, such as rotation on other axes. Another interaction that could be added is the ability to lock the replica in place to prevent accidental movement, or to reset the replica to the original position. Users could also be able to hide the replica from sight, to focus on the 3D model. 

    Other interactions include the ability to show the user's gaze direction, to help other users understand what the user is looking at. This could be done by showing a ray from the user's head to the object they are looking at. With this interaction, users could even be able to create points of interest by looking at an object and performing a gesture. Another possibility is to allow users in the same table to communicate with each other through audio, to further enhance collaboration.

% BETTER TESTING, especially on the collaborative task
    % additional metric: finger movement on transform vs finger movement on balloon selection
    % more careful planning of the tasks and hardware setup
    % test with more users
    % tasks to test the usefullness of teleportation
    % replica scale metric was wrongly calculated